---
title: "Are Better Educated Americans More Likely to Have Extramarital Sex?"
author: "By YUJI SHEN, [yuji.im](http://yuji.im) | APRIL 15, 2015"
output:
  html_document:
    css: ~/Workspace/Web/markdown.css
---

******

```{r knitr.setting,echo=FALSE}
options(width=120)
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      echo = FALSE,
                      fig.align='center')
```

```{r,echo=FALSE,message=F}
setwd('/Users/Yuji/Workspace/R/Duke Statistics')
load("statistics-project-gss.Rdata")
load('inference.Rdata')
library(ggplot2)
library(scales)
library(dplyr)
library(RColorBrewer)
library(reshape2)
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;*This project is based on my final project of course [Data Analysis and Statistical Inference](https://www.coursera.org/course/statistics). Original project only contained one inference, __Chi-squared Test of Independence__. Under the instruction of Josiah and after several revisions, I expand it to include another test, __Z Test of Difference of Two Proportions__, and more visualizations, line charts and bar charts.*

## 1. Introduction
Americans' views on sex-related issues have changed dramatically over the past forty years - more liberal on premarital sex and homosexual sex, but more against extramarital sex, slightly or no much change on teenage sex. Better educated tend to give more accepting on all these topics, but does this mean they have higher probabilities to do these things, or they just have more tolerance? In this study, I will focus on extramarital sex, or extramarital affairs, performing some statistical methods and visualization technique to discuss this question.

## 2. Data
**Description**

Most of my inferences and graphs are based on [General Social Survey Cumulative File, 1972-2012 Coursera Extract ^1^](https://d396qusza40orc.cloudfront.net/statistics%2Fproject%2Fgss1.html). The General Social Survey (GSS) has been monitoring societal change and studying the growing complexity of American society since 1972. You can find the years  of the surveys are performed on the line graph below - almost once a year from 1972 to 1994, then changed to be biennial and keep this routine until now. GSS data is suitable for doing strict statistical inference because it was collected by mainly using stratifying sampling, which is a method of random sampling and will guarantee generalizability of this study. The size of the samples are reasonable, around 2000 observations in each survey.

**Variables**
```{r}
temp <- gss[,c("premarsx", "xmarsex", "teensex", "homosex")]
table <- NULL
for(i in 1:4){
  table <- cbind(table, table(temp[[i]], exclude = NULL))
}
colnames(table) <- c("Premarital Sex", "Extramarital Sex", "Teenage Sex", "Homosexual Sex")
knitr::kable(table, align = "c", caption = "Sexual Behaviors Related Variables Summary")

gss$xmarsex <- factor(gss$xmarsex)
gss$premarsx <- factor(gss$premarsx)
gss$teensex <- factor(gss$teensex)
gss$homosex <- factor(gss$homosex)
```

The sex related variables I will use are:  [PREMARSX](https://d396qusza40orc.cloudfront.net/statistics%2Fproject%2Fgss1.html#premarsx) (Premarital Sex), [XMARSEX](https://d396qusza40orc.cloudfront.net/statistics%2Fproject%2Fgss1.html#xmarsex) (Extramarital Sex), [TEENSEX](https://d396qusza40orc.cloudfront.net/statistics%2Fproject%2Fgss1.html#teensex) (Teenage Sex), [HOMOSEX](https://d396qusza40orc.cloudfront.net/statistics%2Fproject%2Fgss1.html#homosex) (Homosexual Sex). All of them have five exclusive options which represent the opinion of respondents to these behaviors. The "Other" option were only chosen by 82 people in HOMOSEX, which is a relatively small amount, so I omit them for simplifying computation and visualization.

## 3. Trends over Time
The levels in all sex-related questions are the same and ordered, so I encode them into numbers and this makes it much easier to visualize the trends. In the following plot, *__the higher score, the more liberty of society for this behavior__*.

```{r,fig.height=4.5,fig.align='left'}
score <- gss %>%
  filter(homosex != "Ohter") %>%
  group_by(year) %>%
  summarise(xmarsex = mean(xmarsex, na.rm = T),
            premarsx = mean(premarsx, na.rm = T),
            teensex = mean(teensex, na.rm = T),
            homosex = mean(homosex, na.rm = T))

score.melt <- melt(score, id = "year")
score.melt <- transform(score.melt, Issues = factor(variable, levels = c("premarsx", "teensex", "homosex", "xmarsex"), ordered = T))
b <- c(1,2,3,4)

year <- as.Date(paste(unique(gss$year),"-01-01",sep=""))
score.melt$year <- as.Date(paste(score.melt$year,"-01-01",sep=""))

ggplot(score.melt, aes(x = year, y = value, color = Issues)) + geom_line(data =  score.melt[!is.na(score.melt$value), ], size = 2) + xlab("Year") + ylab("Annual Score") + scale_y_continuous(limits = c(1.1,3.1), breaks = c(seq(1,3.5,0.5)), labels = c("Always Wrong\n1.0","1.5","Almst Always Wrg\n2.0", "2.5","Sometimes Wrong\n3.0", "3.5")) + ggtitle("Annual Scores of Attitude to Sex Issues")+scale_color_discrete(labels=c("Premarital Sex","Teenage Sex","Homosexual Sex", "Extramarital Sex")) + scale_x_date(breaks = year, labels = date_format("%Y")) + theme(panel.grid.minor.x = element_blank(), axis.text.x = element_text(angle = 90, vjust = 0.5), legend.position = "bottom", axis.title.x = element_blank())
```

* The most dramatic change is on Homosexual Sex. People change their attitude almost by one degree, which means the liberal level upgraded!
* Another ascender is Premarital Sex. Its change is not as much as the former, but rising steadily and liberal level keeps highest among all these issues.
* Extramarital sex is the ___only one drops over decades___. It is not very obvious, but I will prove that this decrease is statistical significant.
* Teenage Sex did not change too much, and this question only has data since 1988.

## 4. More Educated More Liberal
```{r}
sex <- melt(gss[,c("degree", "xmarsex", "premarsx", "teensex", "homosex")], id = "degree")
sex.prop <- sex %>%
  filter(!is.na(value) & !is.na(degree) & value != "Other") %>%
  group_by(degree, variable, value) %>%
  summarise (n = n()) %>%
  mutate(freq = n / sum(n))

levels(sex.prop$variable) <- c("Extramarital Sex", "Premarital Sex", "Teenange Sex", "Homosexual Sex")

ggplot(sex.prop[sex.prop$value == "Always Wrong", ], aes(x = degree, y = freq, fill = degree)) + geom_bar(stat = "identity", position = position_dodge(), color = "black", width = 0.7) + facet_grid(.~variable) + scale_fill_brewer(palette = "YlOrRd") + ggtitle("'Always Wrong' Proportions for Different Sex Ethics\namong Different Degree (1972-2012)") + scale_y_continuous(label = percent) + theme(axis.text.x = element_blank(), axis.title =element_blank(), axis.ticks.x=element_blank(), legend.position = "bottom", legend.title = element_blank())
```

Besides the trends over time, here are also trends on the education degree. The people received a higher education will have a lower proportion of thinking these sexual ethics as "Always Wrong". This plot seems to convince us that the education will make people become liberal. Is this correct? Next, I will focus on extramarital sex to prove that 

1. There is a relationship between opinion on extramarital sex and education degree.
2. This lower proportion is significant on statistics.

```{r,echo=FALSE,fig.height=4,fig.align="left"}
xmarsex.degree <- gss[ ,c("xmarsex", "degree")]

ggplot(na.omit(xmarsex.degree), aes(x = degree, fill = xmarsex)) + geom_bar(position = "fill", width = 0.7)+ scale_fill_manual(values=c("Not Wrong At All"="#cbc9e2","Sometimes Wrong"="#9e9ac8","Almst Always Wrg"="#756bb1","Always Wrong"="#54278f"), name = "Extramarital\nSex") + ggtitle("Aggregated Proportional Distribution on Extramarital Sex\nby Different Education Degrees (1972-2012)") + coord_flip() +  scale_y_continuous(labels = percent) + theme(axis.title = element_blank(), legend.position = "bottom")
```

### Part 1: Does There Appear to Be a Relationship between Education Degree and Opinion on Extramarital Sex?

**Step 1: Determine Method**

This is a test for independence between two categorical variables (each has more than two levels), so **_Chi-squared Test for Independence_** will be applied.

**Step 2: State Hypothesis**

* ___H<sub>0</sub>___: Education degree and attitude to extramarital sex are independent. 
* ___H<sub>A</sub>___: Education degree and attitude to extramarital sex are dependent.

**Step 3: Check Conditions**

* ___Independence___: This is a random sampling survey with 33397 records after omitting missing values which are less than 10% of population of U.S. Also, each respondent can only choose one answer. The conditions of independence are satisfied.
* ___Sample Size___: The least number in the summary table is 52 (see output below), so each particular scenario have more than five cases. The condition of sample size is satisfied.

**Step 4: Perform Inference**
```{r}
inference(y = xmarsex.degree$xmarsex, x = xmarsex.degree$degree, est = "proportion", type = "ht", method = "theoretical",alternative = 'greater', eda_plot = F, sum_stats = T, inf_plot = F)
```
`inference` function is provided by Duke University's course "Data Analysis and Statistical Inference". ([Source](http://bit.ly/dasi_inference))

**Step 5: Interpret Results**

The p-value is so small, so we can conclude that education degree and the opinion on extramarital sex are not independent among Americans. This means the distribution of four options differs among people with different education degrees. But this is only the first step. Chi-squared test will not tell us what kind of relationship between these two variables. That's why we need Z test.

### Part 2: Is the Bachelors' Proportion Of "Always Wrong" Significantly Greater than Graduates'?

**Step 1: Determine Method**

If the graduate's and the bachelor's proportions are neck to neck, their difference should be around 0. So we will perform a hypothesis test ___(Z Test) on the difference of these two proportions___. This test requires there are only two levels in the response variable, so I aggregate all levels except "Always Wrong" into a new level: "Not Always Wrong", then perform the inference.

**Step 2: State Hypothesis**

* ___H<sub>0</sub>___: Proportion of "Always Wrong" in Bachelors - Proportion of "Always Wrong" in Graduates = 0
* ___H<sub>A</sub>___: Proportion of "Always Wrong" in Bachelors - Proportion of "Always Wrong" in Graduates > 0

**Step 3: Check Conditions**

* ___Independence___
    * ___Within Groups___: This is a random sampling survey with 6947 records after omitting missing values which are less than 10% of the population of U.S. So the sampled observations are independent within people with a different education degree.
    * ___Between Groups___: There is no reason to expect sampled graduates and bachelors to be dependent, so two groups are independent with each other.
* ___Sample Size / Skew___: Number of expected successes and failures are greater than 10 for both groups (see output below), so we can assume that the sampling distribution of the difference between two proportions is nearly normal.

**Step 4: Perform Inference**
```{r}
xmarsex.ci <- na.omit(subset(gss, degree == "Graduate" | degree == "Bachelor")[,c("xmarsex", "degree")])
xmarsex.ci$degree <- factor(xmarsex.ci$degree)
levels(xmarsex.ci$xmarsex) <- c(levels(xmarsex.ci$xmarsex), "Not Always Wrong")
xmarsex.ci[xmarsex.ci$xmarsex != "Always Wrong", ]$xmarsex <- "Not Always Wrong"
xmarsex.ci$xmarsex <- factor(xmarsex.ci$xmarsex)
inference(xmarsex.ci$xmarsex, xmarsex.ci$degree, est = "proportion", type = "ht", method = "theoretical", success = "Always Wrong", alternative = "greater", null = 0, eda_plot = F, inf_plot=F, sum_stats = T)
```
**Step 5: Interpret Results**

A 0 p-value implies that the proportion of "Always Wrong" in Bachelors is significantly greater than the  proportion of "Always Wrong" in Graduates. Thus, we can infer that graduates give more accepting for extramarital sex.

But this is only a test for two groups in education, then I applied the same test on the rest groups following their orders. The result shows that except "Junior College" and "High School" have a 0.2 p-value, all relative lower education degree groups have significant higher proportions of "Always Wrong". Therefore, to some degree, we can say *__better educated Americans are more liberal for extramarital sex__*.

## 5. Liberal ≠ Unruly

However, being liberal about this issue does not equal to the actual performance of people. So the reasonable one would be using the data about the behavior of people's out of marriage affairs. There really is one - [Fair's Extramarital Affairs Data (Table F17.2) ^2^](http://pages.stern.nyu.edu/~wgreene/Text/Edition7/tablelist7.htm). 

I will use the second survey in Fair's Affairs Data, which is conducted by Redbook magazine in 1974 and female  respondents only. A questionnaire on sex was published in the October issue of Redbook, and readers were asked to mail in their answers. Although it was not random sampled, what I am concerned is the distribution of extramarital sex within each group. So this bias seems not a big issue here.

```{r,fig.height=3,fig.align='left'}
affair <- read.csv("TableF17-2.csv")
affair$degree <- factor(affair$V6, labels = c("Grade School", "High School", "Some College", "College Graduate", "Some Graduate School", "Advanced Degree"))
affair$xmarsex <- factor(ifelse(affair$YRB > 0, "Yes", "No"))
affair.table <- prop.table(table(affair$degree, affair$xmarsex), 1)

data <- data.frame(degree = factor(row.names(affair.table), levels = levels(affair$degree)), xmarsex = as.numeric(affair.table[,1]), stringsAsFactors = F)

ggplot(data = data, aes(x = degree, y = xmarsex, fill = degree)) + geom_bar(stat = "identity", color = "black", width = 0.7) + scale_fill_brewer(palette = "YlOrRd", guide = F) + coord_flip() + ylab("") + xlab("") + ggtitle("Proportions of Different Education Degree People\nNever Had Extramarital Sex (1974)") + scale_y_continuous(limits = c(0, 0.8), breaks = seq(0, 0.8, 0.1), labels = percent) + theme(axis.title = element_blank())
```

A REVERSED result! People with more advanced degree have a relative higher proportion for never had extramarital sex! Although better educated accept affairs more, they appear to be much conservative, or say serious about their real actions in lives. I will not use statistical test here because this data do not have generalizability. Here is another graph from GSS can show that how better educated people value their marriage more. 

```{r}
marital <- gss %>%
  filter(!is.na(marital) & !is.na(degree)) %>%
  group_by(year, degree, marital) %>%
  summarise (n = n()) %>%
  mutate(freq = n / sum(n))

marital$year <- as.Date(paste(marital$year,"-01-01",sep=""))

ggplot(subset(marital, marital == "Married"), aes(x = year, y = freq, col = degree)) + geom_line(size = 1.5) + theme(axis.title = element_blank(), legend.title = element_blank(), legend.position = "bottom", panel.grid.minor.x = element_blank(), axis.text.x = element_text(angle = 90, vjust = 0.5)) + scale_y_continuous(label = percent) + scale_color_manual(values = c(brewer.pal(6, "YlOrRd"))) + ggtitle("Marriage Rate by Education Degree (1972 - 2012)") + scale_x_date(breaks = year, labels = date_format("%Y"))
```

This graph describes the marriage rate among all GSS respondents grouped by their degrees for 40 decades. We can see that in the left side of plot, all groups began from similar and high marriage rates, around 80%. As time goes by, all of them dropped due to the change of view on marriage. A fair interesting phenomenon is that the decline speeds for each group are associated with their degree: **The higher education level, the slower decreasing speed on marriage rate**. 

## 6. Conclusion

I came a long way to this conclusion part. After reviewing the trend of sexual viewpoints over four decades, I found that the extramarital sex is the only one given more against by Americans. After comparing the opinion over different education degrees, it seems that better educated are more liberal for these ethics issues. Then, I focus on extramarital sex and use statistical inferences to prove this is the case for whole Americans. Lastly, I use a more direct data about affairs to show actually the better educated are more disciplined in their real lives, and the marriage rate in GSS also implies that people with higher education tend to have a higher chance to keep their marriages over time.

__Receiving higher education is likely to make a people more liberal and tolerant for these sexual ethics, and it also taught him to rule himself and treasure his family in his life.__

### References

1. *Smith, Tom W., Michael Hout, and Peter V. Marsden. General Social Survey, 1972-2012 [Cumulative File]. ICPSR34802-v1. Storrs, CT: Roper Center for Public Opinion Research, University of Connecticut /Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributors], 2013-09-11.  doi:10.3886/ICPSR34802.v1. Source: http://d396qusza40orc.cloudfront.net/statistics/project/gss.Rdata.*
2. *Fair, Redbook Survey on Extramarital Affairs, 6,366 Observations. Source:  http://www.stern.nyu.edu/~wgreene/Text/Edition7/TableF17-1.csv.*
